Live coding environment for a data exploration language

### 2019-02-11 wip-live-programming/paper.tex
Conclusions and space tweaks

As discussed earlier, our implementation updates $\Delta$ during the recursive binding process and
so a stronger version of the property holds -- namely, previews calculated over a graph obtained
directly for the original program $p$ are the same as the values of the fully evaluated program.
We note that this is the case, but do not show it formally to aid the clarity of our formalisation.

The second important property that guarantees that our mechanism always displays correct previews
is determinacy. This makes it possible to cache the previews evaluated using~$\Downarrow$ using
the corresponding graph node as a lookup key.

@@

The mechanism used for caching previews, as discussed in Section~\ref{sec:previews-cache},
keeps a preview or a partial preview $d$ in a lookup table indexed by nodes $\bndclr{v}$. The
Theorem~\ref{thm:determinacy} guarantees that this is a valid strategy. As we update dependency
graph during code editing, previous nodes will continue representing the same sub-expressions.

@@

\begin{figure}[t]
\raggedright
\vspace{0.5em}
\hspace{1.5em}{\small\sffamily Edit contexts of expressions}
\begin{equation*}
\begin{array}{lcl}
K_e[-] &=& K_e[-].m(e_1, \ldots, e_n) ~\lsep~ e.m(e_1, \ldots, e_{l-1}, K_e[-], e_{l+1},\ldots , e_n) ~\lsep~ -\\
K_c[-] &=& \kvd{let}~x = K_e[-] ~\lsep~ K_e[-]\\
\end{array}
\end{equation*}

@@

\caption{Code edit operations that preserve previously evaluated preview}
\label{fig:operations}
\end{figure}

@@

We use the notation $\preview{e}$ to mark the part of the expressions before and after the edit
that have the same value that will not need to be recomputed after the change and we write
$\overline{c}$ and $\overline{e}$ for a list of commands and expressions, respectively.
In some of the edit operations, we also specify an intermediate program that is, in some cases,
semantically different and only has a partial preview. This illustrates a typical way of
working with code in a text editor using cut and paste operations. For example, in
\rname{let-intro-ins}, we assume that the analyst cuts a sub-expression $e$, replaces it with
a variable $x$ and then adds a let binding for a variable $x$ and inserts the expression $e$
from the clipboard. Edit with the same result is captured by \rname{let-intro-del} but here
the analyst first inserts the let binding and then replaces the sub-expression $e$ with a
variable $x$.

@@

\begin{figure}
\raggedright

@@

\begin{figure}[b]
\vspace{-0.5em}
\includegraphics[scale=0.425]{drawing.png}
\caption{Time required to recompute the results of a sample program after individual tokens of
  the program are added or modified for three different evaluations strategies.}
\label{fig:drawing}
\vspace{-0.5em}
\end{figure}

@@

\begin{figure}
\noindent
\includegraphics[scale=0.48]{hist-all.png}
\includegraphics[scale=0.48]{hist-slow.png}
\caption{Histogram showing the distribution of delays incurred when updating previews.
  We show a histogram computed from all delays (left) and only from delays larger
  than 15ms (right).}
\label{fig:drawing-hist}
\end{figure}

@@

\subsubsection{Summary of results}
The purpose of our empirical is not to exactly assess the overhead of our implementation
of the live evaluation strategy algorithm. Instead, our goal is to
illustrate how often can previously evaluated results be reused and the impact this has when
writing code. The experiment presented in this section is small-scale, but it is sufficient
to illustrate the main point.

When recomputing results after every edit operation
using the \emph{call-by-value} strategy, the time needed to update the results grows
continually. The \emph{lazy} strategy removes the overhead for programs that fail to evaluate, but
keeps the same trend. Using our \emph{live} strategy, the computation can reuse values computed
previously and, as a result, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
are significantly faster, because they do not need to recompute operations done previously when
writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
other two strategies.

@@

\begin{figure}[t]
\noindent
\hspace{-1em}
\includegraphics[scale=0.245]{gallery1.png}
\includegraphics[scale=0.245]{gallery2.png}
\caption{\small Data analysis counting the number of libraries per county in the UK. We load
  and aggregate data (left) and then create a chart with a label (right).
  The example is adapted from \url{http://gallery.thegamma.net/73/}.}
\label{fig:gallery}
\end{figure}

@@

\begin{itemize}[itemsep=3pt]
\item The left screenshot uses a type provider for data aggregation introduced by \citet{gamma}
  and briefly discussed in Section~\ref{sec:background-gamma}. For the purpose of
  this paper, the type provider is an external library (with objects, members and reduction relation).
  Using the type provider relies on editor auto-complete which is based on type information. To support
  this, we implement type checking, as discussed in Section~\ref{sec:extra-types}, reusing the mechanism
  of live preview evaluation.

\item The live preview for code written using the data aggregation type provider (left) is shown
  in a panel below the current command. This shows the preview for a sub-expression based on the
  position of the cursor (i.e.~raw data before grouping in the screenshot). In addition to the
  preview, our environment also generates tabs that let the user navigate to individual steps of
  the data transformation for code using the data aggregation type provider.

\item The right screenshot uses a charting library, which is another external library available in
  the environment. Unlike for data aggregation (left), live previews for charts (right) are not
  handled in a special way and we show the result of evaluating the whole command. The screenshot
  shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
  strategy, this is done efficiently without reevaluating the data transformation.
\end{itemize}

\noindent
Conducting an extensive user study that would evaluate the usability of the system from a
human-computer interaction perspective is left for future work. To give anecdotal evidence of
the usability, the example shown in Figure~\ref{fig:gallery} is an analysis developed by an
attendee of our tutorial at Mozilla Festival 2017 who had no prior programming experience.

@@

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\hspace{-3.5em}
\begin{array}{l}
\inference[\rname{mem}~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
  \bndclr{v_0} \vdash \{.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..\} & \bndclr{v_i} \vdash \tau_i }
  { \bnd{mem}(m, s) \vdash \tau }
\\[1.5em]
\inference[\rname{var}~]
  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
  { \bnd{var}(x, s) \vdash \tau' }
\\[1.5em]
\inference[\rname{fun}~]
  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E \\
    \bndclr{v_c} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} &
    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Rules that define type checking of terms and expressions over a dependency graph $(V, E)$}
\label{fig:tc}
\vspace{-0.5em}
\end{figure}

@@

\subsubsection*{Semantics and partial evaluation.}
The evaluation of previews can be seen as a form of partial evaluation \cite{partial}, done in a
way that allows reuse of results. This can be done implicitly or explicitly in the form of
multi-stage programming \cite{metaml}. Both can provide useful perspective for formally analysing
how previews are evaluated. Semantically, the evaluation of previews can be seen as a modality
\cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt}, which,
in turn, can be understood in terms of comonads \cite{cmtt-denotation}. This provides an intriguing
direction for rigorous analysis of the presented system.

% ==================================================================================================
%
% \section{TODO}
%
% TBD: Define live preview somewhere
%
% TBD: In section about dependency graphs, include a subsection discussing how this relates
% to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
%
%
% \cite{selfadjusting}
% \cite{incremental}
%
% Incremental computation
%
%   Self-adjusting computation
%   http://www.cs.cmu.edu/~rwh/theses/acar.pdf
%
%   Incremental Computation with Names
%   https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
%   -> dsl interpreter
%
%   (and other adapton stuff http://adapton.org/)
%
% Type checking
%
%   co-contextual
%   http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
%   http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
%
% https://dl.acm.org/citation.cfm?id=664109
%
\section{Summary}
We present The Gamma, a live coding environment for data exploration. The environment
bridges the gap between spreadsheets and scripting -- live previews give users rapid feedback,
while the final result is a fully reproducible script.

In this paper, we focus on the challenge of efficiently providing live previews and type checking
code during editing in a free-form text editor. This is a challenge, because users can perform
arbitrary text transformations and we cannot recompute previews after each edit.

The key trick is to separate the process into fast \emph{binding phase},
which constructs a dependency graph and slower \emph{evaluation phase} and \emph{type checking phase}
that can cache results, using the nodes from the dependency graph created during binding as keys.
This makes it possible to quickly parse updated code, reconstruct dependency graph and compute
preview using previous, partially evaluated, results.

We describe our approach formally, which serves two purposes. First, we aim to provide easy to
use foundations for the growing and important trend of text-based live coding environments.
Second, we explore the properties of our system and prove that our method does not recompute
previews in a number of common cases and, at the same time, the optimisation still produces
correct previews.

>>>>

\noindent
Our implementation updates $\Delta$ during the recursive binding process and so a stronger version
of the property holds: previews calculated over a graph obtained directly for the original program
$p$ are the same as the values of the fully evaluated program. Our formalisation omits this for
simplicity.

The second important property is determinacy, which makes it possible to cache the previews
evaluated using~$\Downarrow$ using the corresponding graph node as a lookup key.

@@

\noindent
The mechanism used for caching previews (Section~\ref{sec:previews-cache}),
keeps a preview preview $d$ in a lookup table indexed by nodes $\bndclr{v}$.
Theorem~\ref{thm:determinacy} guarantees that this is a valid -- as we update dependency
graph during code editing, previous nodes will continue representing the same sub-expressions.

@@

\begin{figure}[t]
\raggedright
\vspace{-0.5em}
\hspace{1.5em}{\small\sffamily Edit contexts of expressions}
\begin{equation*}
\begin{array}{lcl}
K_e[-] &=& K_e[-].m(e_1, \ldots, e_n) ~\lsep~ e.m(e_1, \ldots, e_{l-1}, K_e[-], e_{l+1},\ldots , e_n) ~\lsep~ -\\
K_c[-] &=& \kvd{let}~x = K_e[-] ~\lsep~ K_e[-]\\
\end{array}
\end{equation*}

@@

\caption{Code edit operations that preserve previously evaluated preview}
\label{fig:operations}
\vspace{-0.5em}
\end{figure}

@@

We use the notation $\preview{e}$ to mark the part of the expressions before and after the edit
that have the same value and will not need to be recomputed after the change; we write
$\overline{c}$ and $\overline{e}$ for a list of commands and expressions, respectively.
In some of the edit operations, we also specify an intermediate program that may be
semantically different and only has a partial preview. This illustrates a typical way of
working with code in a text editor using cut and paste. For example, in
\rname{let-intro-ins}, we the analyst cuts a sub-expression $e$, replaces it with
a variable $x$ and then adds a let binding for a variable $x$ and inserts the expression $e$
from the clipboard. The \rname{let-intro-del} operation captures the same edit, but
done in a different order.

@@

\begin{figure}[b]
\raggedright

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[scale=0.425]{drawing.png}
\caption{Time required to recompute the results of a sample program after individual tokens of
  the program are added or modified for three different evaluations strategies.}
\label{fig:drawing}
\vspace{-0.5em}
\end{figure}

@@

\begin{figure}
\vspace{-0.5em}
\noindent
\includegraphics[scale=0.48]{hist-all.png}
\includegraphics[scale=0.48]{hist-slow.png}
\vspace{-0.25em}
\caption{Histogram showing the distribution of delays incurred when updating previews.
  We show a histogram computed from all delays (left) and only from delays larger
  than 15ms (right).}
\label{fig:drawing-hist}
\vspace{-0.75em}
\end{figure}

@@

\subsubsection{Summary of results}
The purpose of our experiment is not to exactly assess the overhead of our implementation.
Our goal is to illustrate how often can previously evaluated results be reused and the impact
this has when writing code. The experiment presented in this section is small-scale, but it is
sufficient for this purpose. When recomputing results after every edit
using the \emph{call-by-value} strategy, the time needed to update results grows
continually. The \emph{lazy} strategy removes the overhead for programs that fail, but
keeps the same trend. Our \emph{live} strategy reuses values computed
previously. Consequently, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
are significantly faster, because they do not need to recompute operations done previously when
writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
other two strategies.

@@

\begin{figure}[t]
\noindent
\hspace{-1em}
\includegraphics[scale=0.245]{gallery1.png}
\includegraphics[scale=0.245]{gallery2.png}
\caption{\small Data analysis counting the number of libraries per county in the UK. We load
  and aggregate data (left) and then create a chart with a label (right).
  The example is adapted from \url{http://gallery.thegamma.net/73/}.}
\label{fig:gallery}
\vspace{-1em}
\end{figure}

@@

\begin{itemize}[itemsep=3pt]
\item The left screenshot uses a type provider for data aggregation \cite{gamma}
  (Section~\ref{sec:background-gamma}). Our system views the type provider as an external library
  (with objects, members and reduction relation). Type providers rely on type information to
  provide editor auto-complete, which we support by implementing type checking over a
  dependency graph as discussed in Section~\ref{sec:extra-types}.

\item When displaying live preview for code written using the data aggregation type provider (left),
  our environment also generates tabes that show individual steps of the data transformation.
  A tab is selected based on the cursor position and shows preview for the current sub-expression
  (e.g.~raw data before grouping in the screenshot).

\item Another external library provides support for charting (right). The environment does not
  handle charts in a special way and displays the result of evaluating the whole command. The
  screenshot shows a case where the user modifies parameters of the chart. Thanks to our live evaluation
  strategy, this is done efficiently without reevaluating the data transformation.
\end{itemize}

\noindent
A user study to evaluate the usability of the system from a human-computer interaction perspective
is left for future work. As an anecdotal evidence of the usability, the code in
Figure~\ref{fig:gallery} was developed by an attendee of a tutorial at Mozilla Festival 2017
who had no prior programming experience.

@@

\begin{figure}[b]
\vspace{-0.5em}
\begin{equation*}
\hspace{-3.5em}
\begin{array}{l}
\inference[\rname{mem}~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
  \bndclr{v_0} \vdash \{.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..\} & \bndclr{v_i} \vdash \tau_i }
  { \bnd{mem}(m, s) \vdash \tau }
\\[1.5em]
\inference[\rname{var}~]
  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
  { \bnd{var}(x, s) \vdash \tau' }
\\[1.5em]
\inference[\rname{fun}~]
  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E \\
    \bndclr{v_c} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} &
    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Rules that define type checking of terms and expressions over a dependency graph $(V, E)$}
\label{fig:tc}
\vspace{-0.5em}
\end{figure}

@@

\subsubsection*{Semantics and partial evaluation.}
The evaluation of previews is a form of partial evaluation \cite{partial}, done in a
way that allows caching. This can be done implicitly or explicitly in the form of
multi-stage programming \cite{metaml}. Semantically, the evaluation of previews can be seen as a
modality \cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt},
formally modelled using comonads \cite{cmtt-denotation}. This suggests a direction for rigorous
analysis of the presented system.

% ==================================================================================================
%
% \section{TODO}
%
% TBD: Define live preview somewhere
%
% TBD: In section about dependency graphs, include a subsection discussing how this relates
% to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
%
%
% \cite{selfadjusting}
% \cite{incremental}
%
% Incremental computation
%
%   Self-adjusting computation
%   http://www.cs.cmu.edu/~rwh/theses/acar.pdf
%
%   Incremental Computation with Names
%   https://www.cs.tufts.edu/~jfoster/papers/oopsla15.pdf
%   -> dsl interpreter
%
%   (and other adapton stuff http://adapton.org/)
%
% Type checking
%
%   co-contextual
%   http://www.informatik.uni-marburg.de/~seba/publications/cocontextual-type-checking.pdf
%   http://drops.dagstuhl.de/opus/volltexte/2017/7262/pdf/LIPIcs-ECOOP-2017-18.pdf
%
% https://dl.acm.org/citation.cfm?id=664109
%
\section{Summary}
One of the reasons that make spreadsheets easier to use than programming tools is that they
provide instant feedback. We aim to make programming tools as instantaneous as spreadsheets.
We formalized a number of key aspects of simple data analyses such as those don by journalists
and then used our observations to build both theory and simple practical data analytics tools.

Our \emph{data exploration calculus} is a simple formally tractable language
for data exploration. The calculus captures key observations about simple data analyses. They
rely on logic from external libraries, implement few abstractions and are written as lists of
commands. We later revised the calculus to add type checking and simple abstraction mechanism.

Our main technical contribution is a \emph{live preview} mechanism that efficiently evaluates
code during editing and instantly provides a preview of the result. We allow users to edit code
in unconstrained way in an text editor, which makes this particularly challenging. The key trick
is to separate the process into fast \emph{binding phase}, which constructs a dependency graph
and slower \emph{evaluation phase} that can cache results. This makes it possible to quickly parse
updated code, reconstruct dependency graph and compute preview using previous, partially
evaluated, results.

We evaluated our approach in three ways. First, we proved that our mechanism is correct and
that it reuses evaluated values for many common code edit operations. Second, we conducted an
experimental study that illustrates how often are previously evaluated results reused during
typical programming scenario. Thirdly, we used our research as a basis for online data exploration
environment, which shows the practical usability of our work.
